A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-26 model.tex
section on radially varying epsilon

\subsection{Vertical shear}\label{vertshear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p s}{\p z} -\frac{\p
    P}{\p z}\frac{\p s}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\epsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r}\left[\frac{\p\epsilon}{\p z} + \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

>>>>

\subsection{Vertical shear}\label{vertshear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p s}{\p z} -\frac{\p
    P}{\p z}\frac{\p s}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align}\label{vshear2}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\epsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r}\left[\frac{\p\epsilon}{\p z} + \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

### 2016-08-26 result.tex
section on radially varying epsilon

For simplicity we input constant values of the midplane dust-to-gas ratio
$\epsilon_0$ and the characteristic dust thickness $\Hd$. These,
together with the perturbation wavenumber $k$, comprises main the
parameters of the linear problem. We fix the radial power-law index
for the midplane gas density to $p = -1.5$, that for the
temperature profile to $q=-1$; and set the gas disk aspect-ratio
$h_\mathrm{g}=0.05$. These are fiducial values used by 
\citet[][\citetalias{lin15} in this section]{lin15}.

@@

>>>>

For simplicity, we mostly consider constant midplane
dust-to-gas ratios $\epsilon_0$ and its characteristic height
$H_\epsilon$. The latter is specified via the dust layer thickness
$\Hd(r_0)$. (We will allow for radial dependices in $\epsilon_0$ and
$H_\epsilon$ in \S\ref{varHd}.)   
These, together with the perturbation wavenumber $k$, comprises main the
parameters of the linear problem. We fix the radial power-law index
for the midplane gas density to $p = -1.5$, that for the
temperature profile to $q=-1$; and set the gas disk aspect-ratio
$h_\mathrm{g}=0.05$. These are fiducial values used by 
\citet[][\citetalias{lin15} in this section]{lin15}.

@@

\subsection{Effect of a radially-varying dust-to-gas ratio}\label{varHd}
{\bf move to appendix?}

Here we allow the dust-to-gas ratio $\epsilon$ to depend on
$r$. Specifically we let $\epsilon_0\propto r^{-1}$ and
$H_\epsilon \propto \Hg$ (cf. constant values in the previous
calculations). Then
\begin{align*}
  \frac{\p\epsilon}{\p r} =
  \left(\frac{z^2}{H_\epsilon^2}\frac{d\ln{\Hg}}{dr} -
    \frac{1}{r}\right)\epsilon.   
\end{align*}
We fix $Z=0.01$, $\Hd=0.8\Hg$ at $r=r_0$, and
$k\Hg=30$.

Fig. \ref{compare_vshear_varHd} shows that a radially-varying
dust-to-gas ratio increases the magnitude of the vertical shear rate
away from the midplane (see also Eq. \ref{vshear2}). Thus we find
higher VSI growth rates, as shown in Fig. \ref{vsi_dust_varHd}. 
Surfaces mode are more effectively enhanced by the additional vertical 
shear induced by $\p_r\epsilon$. However, all growth rates remain
$O(\hgas\OmK)$. Importantly, the increase in the growth rate of the
fundamental mode is small. Thus we do not expect the radial
dependence in $\epsilon$ to significantly affect the VSI in a dusty
disk.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_Nz2_varHd} 
  \caption{Vertical shear rate (solid) compared to vertical buoyancy
    (dotted) in a locally isothermal, dusty disk 
    with metalicity $Z=0.01$ and dust thickness $\Hd=0.8\Hg$.
    Black curves assume a dust-to-gas ratio that only depends on
    height; whereas the blue curve also allows a radial dependence in
    $\epsilon$. See text for details.  
    \label{compare_vshear_varHd}
    }
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_varHd} 
  \caption{Unstable VSI modes in the disk models of
    Fig. \protect\ref{compare_vshear_varHd}.  
    \label{vsi_dust_varHd}
    }
\end{figure}

### 2016-08-26 summary.tex
general edits
section on radially varying epsilon

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas with minimal modification. In fact, no
modification is needed  
for strictly isothermal or polytropic gas perfectly-coupled to dust:  
the standard adiabatic energy equation substitutes for the dust continuity
equation. When $c_s$ or $K$ is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation. 
This source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15}, which is also
standard in many hydrodynamics codes.

@@

%give numerical example 
We have applied our one-fluid stability analysis to find unstable modes  
similar to that described by \cite{loren15}. We use the same
parameters reported by them as follows. 
%we set $p=q=0$, which eliminates VSI and minimizes streaming
%instabilities as there is no radial pressure gradient at the disk
%midplane.   
The total metalicity is $Z=0.01$ and the dust scale-height $H_\mathrm{d} =
0.4H_\mathrm{g}$, giving a maximum dust-to-gas ratio
$\mathrm{max}(\epsilon) = 0.025$. The midplane stopping time is 
$t_\mathrm{s0}=0.02\OmK^{-1}$. The disk parameters are $p=-1.5,\, q=-1$, and  
$\hgas=0.05$. We set $k = \pi/H_\mathrm{d}$.

Fig. \ref{compare_eigenvals_ts0d02} shows the unstable modes found for
this setup. We find the usual VSI modes, here with $\omega> 3,\, s<
0.2$. However, we also find modes with $\omega<s$. One of these,
marked by the blue asterisk, has 
frequency 
\begin{align*}
\sigma = (0.3631\ii - 0.1251)\hgas\OmK. 
\end{align*}
This growth rate corresponds to $8.8$ orbits, comparable to the $12$
orbits reported by \citeauthor{loren15}. We visualize this mode in
Fig. \ref{result2d_loren}. We find this mode 
persists when $q=0$, so it is not the VSI. 
Using Eq. \ref{streaming_dispersion} with the above parameters,  
we also estimate a much smaller growth 
rate for the streaming instability of 
$O(10^{-3}\hgas\OmK)$, We thus also rule out
SI, as did \citeauthor{loren15}. However, the mode disappears with
$\tstop=0$, so it is associated with finite dust-gas drag.

>>>>

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas with minimal modification. In fact, no
modification is needed  
for strictly isothermal or polytropic gas perfectly-coupled to dust:  
the standard adiabatic energy equation substitutes for the dust continuity
equation. When $c_s$ or $K$ is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation. 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15}, which is also
standard in many hydrodynamics codes.

@@

%give numerical example 
We have applied our one-fluid stability analysis to find unstable modes  
similar to that described by \cite{loren15}. We use the same
parameters reported by them as follows. 
%we set $p=q=0$, which eliminates VSI and minimizes streaming
%instabilities as there is no radial pressure gradient at the disk
%midplane.   
The total metalicity is $Z=0.01$ and the dust scale-height $H_\mathrm{d} =
0.4H_\mathrm{g}$, giving a maximum dust-to-gas ratio
$\mathrm{max}(\epsilon) = 0.025$. The midplane stopping time is 
$t_\mathrm{s0}=0.02\OmK^{-1}$. The gas disk parameters are $(p,q,
\hgas) =(-1.5, -1,0.05)$.  We set $k = \pi/H_\mathrm{d}$.

Fig. \ref{compare_eigenvals_ts0d02} shows the unstable modes found for
this setup. We find the usual VSI modes, here with $\omega> 3\hgas\OmK,\, s<
0.2\hgas\OmK$. However, we also find modes with $\omega<s$. One of these,
marked by the blue asterisk, has 
frequency 
\begin{align*}
\sigma = (0.3631\ii - 0.1251)\hgas\OmK. 
\end{align*}
This growth rate corresponds to $8.8$ orbits, comparable to the $12$
orbits reported by \citeauthor{loren15}. We visualize this mode in
Fig. \ref{result2d_loren}. We find this mode 
persists when $q=0$, so it is not the VSI. 
Using Eq. \ref{streaming_dispersion} with the above parameters,  
we also estimate a much smaller growth 
rate for the streaming instability of 
$O(10^{-3}\hgas\OmK)$, We thus also rule out
SI, as did \citeauthor{loren15}. However, the mode disappears with
$\tstop=0$, so it is associated with finite dust-gas drag.
